---
title: "Dog Breeds"
author: "Fanyi Zeng"
date: "3/16/2022"
output: github_document
---

The data comes from the American Kennel Club.
source: https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-02-01/readme.md

breed_traits - trait information on each dog breed and scores for each trait (wide format)

trait_description - long descriptions of each trait and values corresponding to Trait_Score

breed_rank_all - popularity of dog breeds by AKC registration statistics from 2013-2020

```{r load library, message=FALSE}
library(tidyverse)
library(tidymodels)
library(dsbox)
library("Hmisc")
library(corrplot)
```

```{r load data, message=FALSE}
breed_traits <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-02-01/breed_traits.csv')
trait_description <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-02-01/trait_description.csv')
breed_rank_all <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-02-01/breed_rank.csv')
```

### Breed Ranking

Let's take a look at the breed ranking dataset. It shows the ranking of 195 dog breeds across 8 years (2013-2020).

```{r breedranking}
breed_rank_all
```

First, I convert the data format from wide to long, in order to visualize the data.

```{r rank}
rank_long <- breed_rank_all %>%
  select(-links, -Image) %>%
  pivot_longer(-Breed, "year") %>%
  mutate(value = as.numeric(value))
```

It is important to convert the year into numeric values.

```{r remove}
rank_long <- rank_long %>%
  mutate(year = str_remove(year, " Rank")) %>%
  mutate(year = as.numeric(year))
```

I am interested in how the 2020's top 10 dogs perform across years, so I only keep the top 10 dogs in the original breed ranking dataset.

```{r top10}
top10in2020 <- rank_long %>%
  filter(year==2020 & value <= 10 & value >=1)
top10 <- semi_join(rank_long, top10in2020, "Breed")
```

Now we can visualize the ranking data. 

```{r trend}
top10 %>%
  ggplot(aes(x=year, y=value, group=Breed, color=Breed)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks=seq(2013,2020,1)) +
  scale_y_continuous(breaks=seq(1,13,1)) +
  labs(title="2020's Top 10 Dog Breeds' Ranking Over Time (2013-2020)", subtitle="By year", y="ranking")
```

We can see that Labrador Retrievers have been ranked #1 for 8 years. Similarly, German shepherd dogs, Golden Retrievers, and bulldogs have always been relatively popular. While Beagles have decreased in ranking, poodles and pointers have gained popularity over the years. As the "dark horses", French bulldogs have gone through a spike in ranking, from #11 in 2013 to #2 in 2020. Dachshunds have experienced a decrease then an increase in ranking, resulting in the same ranking in 2020 as in 2013.

### Breed Traits

Now let's take a look at the breed traits dataset.

```{r breedtraits}
colnames(breed_traits)
```

I am curious about the relationships among the traits, so I run correlations between every two traits. I have taken out breed name, because it is irrelevant, as well as coat type and coat length, because they are string variables. The codes below returns a r value matrix and a p value matrix.

```{r cormatrix}
cor<-rcorr(as.matrix(breed_traits[,c(-1,-8,-9)]))
```

The number of variables and matrices makes the results hard to read, so I decide to visualize the results with a colored plot. The larger a correlation, the larger its circle, and the deeper the color of its circle. Blue indicates a positive correlation, and red indicates a negative correlation. Insignificant correlations are left blank.

```{r corrplot}
corrplot(cor$r, type="upper", order="hclust", 
         p.mat = cor$p, sig.level = 0.05, insig = "blank")
```

We can see some strong correlations around the middle and at the bottom. It seems that affectionate with family, adaptability, open to strangers, playfulness, and trainability are strongly correlated with one another, and these four variables are moderately to weakly correlated with good with children, good with other dogs, and barking level. 

Playfulness (enthusiam level even past the puppy years) are highly correlated with trainability (willingness to learn new things), good with young children, energy level, and mental stimulation.

For some reason, drooling level is negatively correlated with barking level and energy level. Breeds that drool a lot don't like to bark or exercise as often.

Watchdog/protective nature is negatively correlated with good with other dogs. "Watchdog" breeds are alert to the surrounding environment and potential threats, and they tend not be welcoming to other dogs, young children, or strangers. These breeds also drool relatively more and require more mental stimulation.

Strangely, shedding level (how much hair they shed) and coat grooming frequency (how much grooming is required) are negatively correlated, which does not make much sense to me. According to the definition of the shedding level variable, high shedding breeds require more grooming.

In summary, there are roughly several "groups" of dog breeds: (the first type and the second type are highly correlated, in fact)

1. the affectionate type, who are very lovey-dovey towards family and everyone they know well; they are adaptable to changing living environment; they are friendly to strangers and other dogs. They are everyone's best buddies.

2. the playful type, who like to play a lot, learn new tricks, be around young children, and need a lot of physical as well as mental exercise on a daily basis. They are bundles of energy.

3. the watchdog type, who are more cautious about the environment, more wary of other dogs and strangers, less patient around young children, drool more, and require more mental stimulation. They are loyal guard dogs.

### Ranking and Traits

Lastly, I would like to examine the relationship between 2020 ranking and traits. The breed traits dataset is already organized in the order of 2020 breed ranking, so I just need to add the 2020 ranking column at the end of the breed traits dataset.

```{r rank&traits}
breed_traits <- breed_traits %>%
  mutate(ranking = breed_rank_all$`2020 Rank`)
```

Then I run a multiple linear regression. It seems that the breed traits explain about 18% variance in 2020's breed ranking. Among them, playfulness, watchdog nature, and barking level are significant predictors, and affectionate with family and drooling level are marginally significant.

```{r model}
m_rank <- lm(ranking ~., data = breed_traits[,c(-1,-8,-9)])
summary(m_rank)
```

By including playfulness, watchdog nature, and barking level, affectionate with family, and drooling level in the new model, affectionate with family becomes non-significant at .09. So I further remove affectionate with family, and the other four variables are significant.

Playfulness is the most significant predictor of ranking. Since playfulness is positively correlated with barking level, the latter becomes significant too. Playfulness is also very strongly correlated with other good characteristics such as affectionate with family, good with other dogs, good with children, etc., so it makes sense that playful dogs are popular.

It is interesting that drooling level is significant, and it seems to be more significant than watchdog nature, which it is positively correlated with. There might be some characteristics that people value in dogs that happen to drool a lot.

```{r model2}
m_rank2 <- lm(ranking ~., data = breed_traits[,c(7,11,12,16,18)])
summary(m_rank2)
```

This is the most interesting dataset I have analyzed so far!
